Overt attentional capture by reward-related stimuli overcomes inhibitory suppression

### 2019-04-29 analysis/paper/paper.tex
fixed typos in paper and readme

For example, across two experiments Gaspelin et al. (2017) had
participants perform variants of a commonly used procedure for examining
attentional capture---the \emph{additional singleton task} (Theeuwes,
1991, 1992)---while recording eye movements. Eye movements provide a
useful index of attention since an eye movement to a location (a
saccade) is necessarily preceded by a shift of spatial attention to that
location (Deubel \& Schneider, 1996). In this task, participants were
required to locate and respond to a shape-singleton target that was
presented in an array of non-target shapes. On half of the trials, one
of the non-target shapes was rendered in a different colour from the
rest of the array so as to be a physically salient colour-singleton
distractor. If participants were \emph{more} likely to make a saccade
towards the singleton distractor than one of the nonsingleton
non-targets, this would indicate that the singleton distractor had
captured attention. By contrast, if participants were \emph{less} likely
to make a saccade to the singleton distractor than one of the
nonsingleton non-targets, this would suggest that attention to the
singleton distractor had been suppressed. In Gaspelin et al. (2017)'s
Experiment 1, participants were encouraged to adopt a strategy of
searching for the \enquote{odd-one-out} in the display (i.e,
\emph{singleton-search mode}; Bacon \& Egeth, 1994), by having the
target be randomly either a circle amongst diamonds or a diamond amongst
circles on each trial (see Figure \ref{fig:methodFig}A for an example
from the current study). Under these conditions, both stimulus-driven
(Theeuwes, 1992) and goal-directed (Bacon \& Egeth, 1994) accounts of
attentional selection would predict capture, as the template that
participants use to locate the target (i.e., any singleton) is broad
enough to include the singleton distractor. As expected, under these
singleton-search conditions Gaspelin et al. found that participants'
overt attention was captured by the salient singleton. In Experiment 2,
participants were instead encouraged to adopt a more specific target
template (i.e., \emph{feature-search mode}; Bacon \& Egeth, 1994), by
consistently mapping the target identity to one shape (e.g., search for
a circle on every trial), and having a heterogeneous set of shapes act
as the non-targets (see Figure \ref{fig:methodFig}B). This manipulation
makes searching for any singleton a highly inefficient strategy, as
there are multiple singleton nontargets in the display (e.g., a circle,
a square, a triangle, and a red shape). Thus, participants should be
encouraged to search for the specific features that define the target
(i.e., the \enquote{circleness} of the circle). Under these conditions,
attention to the salient colour-singleton distractor was suppressed
relative to the other non-targets, suggesting that the visual attention
system treated the salient singleton differently from the other
non-targets (inconsistent with goal-directed accounts), yet did not
capture gaze (inconsistent with stimulus-driven accounts). By contrast,
the signal suppression hypothesis can account for these findings by
suggesting that the singleton generated an attentional priority signal
as a consequence of its physical salience, but this signal could be
suppressed by an inhibitory mechanism under conditions that promoted a
more specific search strategy (i.e., feature search). Importantly, this
suppression effect was observed even for the most rapidly generated eye
movements, which was taken as evidence that the suppression effect
operates early in processing, before attention is initially allocated.

@@

The task was based on that used by Gaspelin et al. (2017) and Pearson et
al. (2015, 2016). Each trial consisted of a fixation display, a search
display, and a feedback display (see Figure \ref{fig:methodFig}C). All
stimuli were presented on a black background. The fixation display
consisted of a white cross (0.5\(^\circ\) visual angle), presented in
the centre of the screen and surrounded by a white circle (diameter
3.0\(^\circ\)). The search display consisted of four filled shapes (size
1.6\(^\circ\) \(\times\) 1.6\(^\circ\)) presented above, below, left and
right of screen centre (eccentricity 4.5\(^\circ\)). One of the shapes
was designated as the target, and the other three shapes were designated
non-targets. For all participants, the target was randomly determined to
be either a circle or a diamond on each trial.\footnote{This aspect of
  our design differs from that of Gaspelin et al. (2017). In their
  study, for half of participants in the feature search condition the
  target was always a circle, and for other participants the target was
  always a diamond. By contrast, for the singleton search condition, for
  all participants the target was a circle among diamonds on some
  trials, and a diamond among circles on others. That is, target
  identity (circle/diamond) was manipulated between-subjects for the
  feature search condition, and within-subjects for the singleton search
  condition. In the current study, we equated the two conditions by
  varying target identity within-subjects for all participants. This
  ensured that any difference between the two search conditions was
  driven by a difference in the specificity of participants' target
  template and not by a difference in participants' history of selecting
  the target shape (e.g., Maljkovic \& Nakayama, 1994).} The identity of
the non-target shapes varied by search condition. For the
\emph{singleton search} condition, the non-target stimuli were all
rendered as the alternative target shape (i.e., if the target was a
diamond, all of the non-targets were circles; Figure
\ref{fig:methodFig}B). For the \emph{feature search} condition, the
non-target stimuli were randomly selected without replacement from a set
of two squares and two hexagons (such that one of the non-targets was
always a shape singleton), and were never rendered as the alternative
target shape (Figure \ref{fig:methodFig}C). The target and two of the
non-targets were always rendered in grey (CIE \(x\), \(y\) chromaticity
coordinates of .327/.400). On the majority of trials, one of the
non-target shapes (the \emph{singleton distractor}) was rendered in
either orange or blue (CIE \(x\), \(y\) coordinates .492/.445 and
.192/.216, respectively). On the remaining trials, all shapes were
rendered in grey such that there was no singleton distractor. The values
of blue and orange had similar luminance (\textasciitilde{}24.5
cd/m\textsuperscript{2}), which was higher than that of grey
(\textasciitilde{}8.3 cd/m\textsuperscript{2}). The feedback display
showed the points earned on the current trial. If the response time (RT)
exceeded the soft-timeout threshold (see below), the message
\enquote{Too slow} appeared below the reward feedback.

@@

\caption{Trial structure of the current task. (A) Participants in
the singleton search condition experienced search displays where the
target was a unique shape (either a circle amongst diamonds or diamond
amongst circles). (B) Participants in the feature search condition
experienced search displays where the target was either a circle or
diamond shape presented amongst a set of heterogenous shapes. On most
trials, one of the non-target shapes was rendered in either orange or
blue so as to be a colour singleton distractor, leaving two other
nonsingleton non-targets in the display. (C) Participants began by
fixating on a centrally presented fixation cross. A search display then
appeared, and participants were required to make a rapid saccade to the
target. The colour of the singleton distractor indicated the available
reward (10 or 500 points), but looking at the distractor led to reward
omission.}\label{fig:methodFig}
\end{figure}

@@

Participants were randomly assigned to either the singleton search or
feature search conditions. For half of the participants in each
condition, orange was the high-reward colour and blue was the low-reward
colour; these colour-reward relationships were reversed for the
remaining participants. There were three types of trial: (1)
\emph{high-reward trials}, in which the singleton distractor was
rendered in the high-reward colour and 500 points were available for a
rapid respnse (see below); (2) \emph{low-reward trials}, in which the
singleton distractor was rendered in the low-reward colour and 10 points
were available; and (3) \emph{distractor-absent trials}, in which all
non-targets were rendered in grey and 10 points were available. The
experiment comprised 8 blocks of 60 trials each, for a total of 480
experimental trials. Each block consisted of 20 high-reward trials, 20
low-reward trials, and 20 distractor-absent trials, in random order.
Participants took a short break between blocks.

@@

ANOVA analysis of the percentage of reward omissions with factors of
trial type (high-reward vs low-reward) and search condition (singleton
search vs feature search) revealed a main effect of trial type
\(F(1, 62) = 43.11\), \(p < .001\), \(\hat{\eta}^2_p = .410\). Planned
paired-samples \emph{t}-tests confirmed that participants triggered more
reward omissions (i.e., were more likely to look at the colour-singleton
distractor) on high-reward trials than low-reward trials in both the
singleton search condition, \(t(31) = 5.87\), \(p < .001\), \(d_z=\)
1.04 (Figure \ref{fig:OmissionPlot}A) and feature search condition,
\(t(31) = 3.96\), \(p < .001\), \(d_z=\) 0.70 (Figure
\ref{fig:OmissionPlot}B), demonstrating an effect of reward on overt
attentional capture (i.e., a VMAC effect) regardless of whether
participants were encouraged to adopt a more specific target template.
Importantly, search condition did not exert a significant main effect,
\(F(1, 62) = 0.21\), \(p = .647\), \(\hat{\eta}^2_p = .003\), or
significantly interact with the effect of trial type,
\(F(1,62) < 0.01\), \(p=.985\), \(\hat{\eta}^2_p < .0001\). In order to
quantify support for the null hypothesis that the influence of reward on
attentional capture did not vary depending on participants' target
template, we conducted a Bayesian \emph{t}-test to compare the magnitude
of the VMAC effect across the two search conditions. This revealed a
Bayes Factor of \(\mathrm{BF}_{\textrm{01}} = 3.91\), indicating
moderate support for the null hypothesis of no difference in the size of
the VMAC effect under singleton search and fearure search conditions.

@@

An inconsistency between the results of the current study and those of
Gaspelin et al. (2017) comes from the time-course of the attentional
capture and suppression effects that were observed. According to the
signal suppression hypothesis, suppression is \emph{proactive}---in that
it operates prior to the initial allocation of attention---and so
suppression should be observed even for the most rapidly generated
saccades. Consistent with this idea, Gaspelin et al. (2017) observed an
overt attentional suppression effect to physically salient singleton
distractors amongst the fastest quartile of saccade latencies. By
contrast, in the current experiment we observed overt attentional
capture in the fastest quartile of saccade latencies, with the
suppression effect emerging on trials with slower saccades.
Methodological differences between Gaspelin et al.\enquote{s (2017)
experiment and the current study may be responsible for this
discrepancy. For example, the availability of reward on low-reward
trials may have influenced capture, either by somewhat increasing the
attentional priority of the stimulus feature associated with low reward,
or by otherwise influencing participants} motivation to respond quickly.
Alternatively, the discrepancy in findings may reflect our use of two
potential targets (circle and diamond) in the feature search condition,
as opposed to the same target on all trials in Gaseplin et al.'s
procedure. Having two possible targets could reduce the influence of
trial-by-trial upweighting of target features (Maljkovic \& Nakayama,
1994), but may also result in less effective feature search (but see
Grubert \& Eimer, 2015; e.g., Houtkamp \& Roelfsema, 2009). Another
possible explanation comes from the use of an intermixed trial
structure: as participants randomly experienced trials containing either
the high- or low-reward colour throughout the course of the task, it is
possible that the proactive down-weighting of the salient-but-irrelevant
features could not build up enough to prevent capture for the most
rapidly initiated saccades (Gaspelin \& Luck, 2019). How these
methodological differences might influence the time-course of overt
attentional capture and suppression remains a question for future
research. Nonetheless, the finding does raise questions about whether
the attentional suppression effect that we have observed is truly the
result of proactive attentional suppression of salient stimuli, as
suggested by the signal suppression hypothesis, or potentially the
result of rapid covert disengagement from the salient stimuli (see
Theeuwes, 2010; van Zoest et al., 2004 for similar arguments).

>>>>

For example, across two experiments Gaspelin et al. (2017) had
participants perform variants of a commonly used procedure for examining
attentional capture---the \emph{additional singleton task} (Theeuwes,
1991, 1992)---while recording eye movements. Eye movements provide a
useful index of attention since an eye movement to a location (a
saccade) is necessarily preceded by a shift of spatial attention to that
location (Deubel \& Schneider, 1996). In this task, participants were
required to locate and respond to a shape-singleton target that was
presented in an array of non-target shapes. On half of the trials, one
of the non-target shapes was rendered in a different colour from the
rest of the array so as to be a physically salient colour-singleton
distractor. If participants were \emph{more} likely to make a saccade
towards the singleton distractor than one of the nonsingleton
non-targets, this would indicate that the singleton distractor had
captured attention. By contrast, if participants were \emph{less} likely
to make a saccade to the singleton distractor than one of the
nonsingleton non-targets, this would suggest that attention to the
singleton distractor had been suppressed. In Gaspelin et al. (2017)'s
Experiment 1, participants were encouraged to adopt a strategy of
searching for the \enquote{odd-one-out} in the display (i.e.,
\emph{singleton-search mode}; Bacon \& Egeth, 1994), by having the
target be randomly either a circle amongst diamonds or a diamond amongst
circles on each trial (see Figure \ref{fig:methodFig}A for an example
from the current study). Under these conditions, both stimulus-driven
(Theeuwes, 1992) and goal-directed (Bacon \& Egeth, 1994) accounts of
attentional selection would predict capture, as the template that
participants use to locate the target (i.e., any singleton) is broad
enough to include the singleton distractor. As expected, under these
singleton-search conditions Gaspelin et al. found that participants'
overt attention was captured by the salient singleton. In Experiment 2,
participants were instead encouraged to adopt a more specific target
template (i.e., \emph{feature-search mode}; Bacon \& Egeth, 1994), by
consistently mapping the target identity to one shape (e.g., search for
a circle on every trial), and having a heterogeneous set of shapes act
as the non-targets (see Figure \ref{fig:methodFig}B). This manipulation
makes searching for any singleton a highly inefficient strategy, as
there are multiple singleton nontargets in the display (e.g., a circle,
a square, a triangle, and a red shape). Thus, participants should be
encouraged to search for the specific features that define the target
(i.e., the \enquote{circleness} of the circle). Under these conditions,
attention to the salient colour-singleton distractor was suppressed
relative to the other non-targets, suggesting that the visual attention
system treated the salient singleton differently from the other
non-targets (inconsistent with goal-directed accounts), yet did not
capture gaze (inconsistent with stimulus-driven accounts). By contrast,
the signal suppression hypothesis can account for these findings by
suggesting that the singleton generated an attentional priority signal
as a consequence of its physical salience, but this signal could be
suppressed by an inhibitory mechanism under conditions that promoted a
more specific search strategy (i.e., feature search). Importantly, this
suppression effect was observed even for the most rapidly generated eye
movements, which was taken as evidence that the suppression effect
operates early in processing, before attention is initially allocated.

@@

The task was based on that used by Gaspelin et al. (2017) and Pearson et
al. (2015, 2016). Each trial consisted of a fixation display, a search
display, and a feedback display (see Figure \ref{fig:methodFig}C). All
stimuli were presented on a black background. The fixation display
consisted of a white cross (0.5\(^\circ\) visual angle), presented in
the centre of the screen and surrounded by a white circle (diameter
3.0\(^\circ\)). The search display consisted of four filled shapes (size
1.6\(^\circ\) \(\times\) 1.6\(^\circ\)) presented above, below, left and
right of screen centre (eccentricity 4.5\(^\circ\)). One of the shapes
was designated as the target, and the other three shapes were designated
non-targets. For all participants, the target was randomly determined to
be either a circle or a diamond on each trial.\footnote{This aspect of
  our design differs from that of Gaspelin et al. (2017). In their
  study, for half of participants in the feature search condition the
  target was always a circle, and for other participants the target was
  always a diamond. By contrast, for the singleton search condition, for
  all participants the target was a circle among diamonds on some
  trials, and a diamond among circles on others. That is, target
  identity (circle/diamond) was manipulated between-subjects for the
  feature search condition, and within-subjects for the singleton search
  condition. In the current study, we equated the two conditions by
  varying target identity within-subjects for all participants. This
  ensured that any difference between the two search conditions was
  driven by a difference in the specificity of participants' target
  template and not by a difference in participants' history of selecting
  the target shape (e.g., Maljkovic \& Nakayama, 1994).} The identity of
the non-target shapes varied by search condition. For the
\emph{singleton search} condition, the non-target stimuli were all
rendered as the alternative target shape (i.e., if the target was a
diamond, all of the non-targets were circles; Figure
\ref{fig:methodFig}A). For the \emph{feature search} condition, the
non-target stimuli were randomly selected without replacement from a set
of two squares and two hexagons (such that one of the non-targets was
always a shape singleton), and were never rendered as the alternative
target shape (Figure \ref{fig:methodFig}B). The target and two of the
non-targets were always rendered in grey (CIE \(x\), \(y\) chromaticity
coordinates of .327/.400). On the majority of trials, one of the
non-target shapes (the \emph{singleton distractor}) was rendered in
either orange or blue (CIE \(x\), \(y\) coordinates .492/.445 and
.192/.216, respectively). On the remaining trials, all shapes were
rendered in grey such that there was no singleton distractor. The values
of blue and orange had similar luminance (\textasciitilde{}24.5
cd/m\textsuperscript{2}), which was higher than that of grey
(\textasciitilde{}8.3 cd/m\textsuperscript{2}). The feedback display
showed the points earned on the current trial. If the response time (RT)
exceeded the soft-timeout threshold (see below), the message
\enquote{Too slow} appeared below the reward feedback.

@@

\caption{Trial structure of the current task. (A) Participants in
the singleton search condition experienced search displays where the
target was a unique shape (either a circle amongst diamonds or diamond
amongst circles). (B) Participants in the feature search condition
experienced search displays where the target was either a circle or
diamond shape presented amongst a set of heterogeneous shapes. On most
trials, one of the non-target shapes was rendered in either orange or
blue so as to be a colour singleton distractor, leaving two other
nonsingleton non-targets in the display. (C) Participants began by
fixating on a centrally presented fixation cross. A search display then
appeared, and participants were required to make a rapid saccade to the
target. The colour of the singleton distractor indicated the available
reward (10 or 500 points), but looking at the distractor led to reward
omission.}\label{fig:methodFig}
\end{figure}

@@

Participants were randomly assigned to either the singleton search or
feature search conditions. For half of the participants in each
condition, orange was the high-reward colour and blue was the low-reward
colour; these colour-reward relationships were reversed for the
remaining participants. There were three types of trial: (1)
\emph{high-reward trials}, in which the singleton distractor was
rendered in the high-reward colour and 500 points were available for a
rapid response (see below); (2) \emph{low-reward trials}, in which the
singleton distractor was rendered in the low-reward colour and 10 points
were available; and (3) \emph{distractor-absent trials}, in which all
non-targets were rendered in grey and 10 points were available. The
experiment comprised 8 blocks of 60 trials each, for a total of 480
experimental trials. Each block consisted of 20 high-reward trials, 20
low-reward trials, and 20 distractor-absent trials, in random order.
Participants took a short break between blocks.

@@

ANOVA analysis of the percentage of reward omissions with factors of
trial type (high-reward vs low-reward) and search condition (singleton
search vs feature search) revealed a main effect of trial type
\(F(1, 62) = 43.11\), \(p < .001\), \(\hat{\eta}^2_p = .410\). Planned
paired-samples \emph{t}-tests confirmed that participants triggered more
reward omissions (i.e., were more likely to look at the colour-singleton
distractor) on high-reward trials than low-reward trials in both the
singleton search condition, \(t(31) = 5.87\), \(p < .001\), \(d_z=\)
1.04 (Figure \ref{fig:OmissionPlot}A) and feature search condition,
\(t(31) = 3.96\), \(p < .001\), \(d_z=\) 0.70 (Figure
\ref{fig:OmissionPlot}B), demonstrating an effect of reward on overt
attentional capture (i.e., a VMAC effect) regardless of whether
participants were encouraged to adopt a more specific target template.
Importantly, search condition did not exert a significant main effect,
\(F(1, 62) = 0.21\), \(p = .647\), \(\hat{\eta}^2_p = .003\), or
significantly interact with the effect of trial type,
\(F(1,62) < 0.01\), \(p=.985\), \(\hat{\eta}^2_p < .0001\). In order to
quantify support for the null hypothesis that the influence of reward on
attentional capture did not vary depending on participants' target
template, we conducted a Bayesian \emph{t}-test to compare the magnitude
of the VMAC effect across the two search conditions. This revealed a
Bayes Factor of \(\mathrm{BF}_{\textrm{01}} = 3.91\), indicating
moderate support for the null hypothesis of no difference in the size of
the VMAC effect under singleton search and feature search conditions.

@@

An inconsistency between the results of the current study and those of
Gaspelin et al. (2017) comes from the time-course of the attentional
capture and suppression effects that were observed. According to the
signal suppression hypothesis, suppression is \emph{proactive}---in that
it operates prior to the initial allocation of attention---and so
suppression should be observed even for the most rapidly generated
saccades. Consistent with this idea, Gaspelin et al. (2017) observed an
overt attentional suppression effect to physically salient singleton
distractors amongst the fastest quartile of saccade latencies. By
contrast, in the current experiment we observed overt attentional
capture in the fastest quartile of saccade latencies, with the
suppression effect emerging on trials with slower saccades.
Methodological differences between Gaspelin et al.'s (2017) experiment
and the current study may be responsible for this discrepancy. For
example, the availability of reward on low-reward trials may have
influenced capture, either by somewhat increasing the attentional
priority of the stimulus feature associated with low reward, or by
otherwise influencing participants' motivation to respond quickly.
Alternatively, the discrepancy in findings may reflect our use of two
potential targets (circle and diamond) in the feature search condition,
as opposed to the same target on all trials in Gaspelin et al.'s
procedure. Having two possible targets could reduce the influence of
trial-by-trial upweighting of target features (Maljkovic \& Nakayama,
1994), but may also result in less effective feature search (but see
Grubert \& Eimer, 2015; e.g., Houtkamp \& Roelfsema, 2009). Another
possible explanation comes from the use of an intermixed trial
structure: as participants randomly experienced trials containing either
the high- or low-reward colour throughout the course of the task, it is
possible that the proactive down-weighting of the salient-but-irrelevant
features could not build up enough to prevent capture for the most
rapidly initiated saccades (Gaspelin \& Luck, 2019). How these
methodological differences might influence the time-course of overt
attentional capture and suppression remains a question for future
research. Nonetheless, the finding does raise questions about whether
the attentional suppression effect that we have observed is truly the
result of proactive attentional suppression of salient stimuli, as
suggested by the signal suppression hypothesis, or potentially the
result of rapid covert disengagement from the salient stimuli (see
Theeuwes, 2010; van Zoest et al., 2004 for similar arguments).
